 

### 2020-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. [@nakamuraForestsTheirCanopies2017, @ozanneBiodiversityMeetsAtmosphere2003]  
With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).

@@

Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a, Senfetal.2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]

@@

We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellweger_forest_2020, @suggittExtinctionRiskClimate2018, @scheffersMicrohabitatsReduceAnimal2013,@defrenneMicroclimateModeratesPlant2013]

@@

(@niinemetsWithincanopyVariationRate2004a highlights biophysical and photosynthetic gradient interactions along vertical canopy profile, however our current knowledge of this topic is lacking.)

@@

Most notably, light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
(Fig. **2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a, Chazdon & Fetcher 1984; Parker et al. 2019]
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984, @tymenQuantifyingMicroenvironmentalVariation2017, @smithSeasonalDroughtrelatedChanges2019]. 
This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.[@fausetDifferencesLeafThermoregulation2018, @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006, @doughtyAreTropicalForests2008]

@@

**Wind speeds are also higher at the top of the canopy (Fig. **2x**).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [e.g., REFS; @mcgregor_tree_nodate] and open savannas (Curtis et al. 2018).
This results in higher boundary layer conductance ($g_{a}$) for canopy leaves. Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata. Large $g_{a}$ for canopy leaves results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990].

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. **2x**).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019, @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019]. 
Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017] 
"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy"
However, similar maximum temperatures have been observed during the dry season in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
Minimum $T_{air}$ is also buffered by forest canopies under some conditions. 
Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate, @nakamuraForestsTheirCanopies2017].
*(Add something about risk of freeze.)*--> [@nakamuraForestsTheirCanopies2017]

@@

Forest canopies moderate relative humidity in below-canopy microclimates along with temperature. In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate. With rising temperatures, this trend in maximum $T_{air} and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012]

RH, in combination with temperature, determines vapor pressure deficit, $VPD$
- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018, Tymen et al. 2017]
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004]

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
-	Higher in understory, particularly at dusk (Koike et al. 2001).
-	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
-Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004]

@@

**The strength of this buffering varies across forests, being influenced by both forest characteristics and the biophysical environment.**
First, buffering increases with canopy cover.
* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015]
*Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017,  @martinCovarianceSunShade2020a]

@@

Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b]. Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@@niinemetsWorldwideAnalysisWithincanopy2015] Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregular and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016, @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al. 2020, REFS)

@@

Photosynthesis increases with increasing temperature and saturates at quatum flux density of 400 to 700nm m-2 s-1, beyond this the leaf experiences heat stress; undergoes non-photochemical quenching, photoinhibition and leaf necrosis. Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves, especially in conditions of water stress or drought. [@niinemetsWorldwideAnalysisWithincanopy2015]

**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5; *revisit this!*) (Campbell and Norman 1998).
This observed by earlier studies suggested that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of Topt, Vmax anD Jmax along the gradient [@niinemetsWithincanopyVariationRate2004a, Harley and Baldocchi 1995, Friend, 2001].
However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photoshythesis in sun and shade leaves.
For 3 species in Panama, (Slot et al. 2019) found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant. Similiar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
Mau et al. found no trend along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].

>>>>

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**

Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. [@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003]  
With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. 
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).

@@

Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019]. Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; Senfetal.2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]

@@

We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellweger_forest_2020; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013; @defrenneMicroclimateModeratesPlant2013]

@@

(@niinemetsWithincanopyVariationRate2004a highlights biophysical and photosynthetic gradient interactions along the vertical canopy profile, however our current knowledge of this topic is lacking.)

@@

Most notably, light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
(Fig. **2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al. 2019]
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. 
This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature [@fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].

@@

**Wind speeds are also higher at the top of the canopy (Fig. **2x**).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [e.g., REFS; @mcgregor_tree_nodate] and open savannas (Curtis et al. 2018).
This results in higher boundary layer conductance ($g_{a}$) for canopy leaves. Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata. Large $g_{a}$ for canopy leaves results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a].

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. **2x**).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019]. 
Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017]. 
"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy".
However, similar maximum temperatures have been observed during the dry season in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
Minimum $T_{air}$ is also buffered by forest canopies under some conditions. 
Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
*(Add something about risk of freeze.)*--> [@nakamuraForestsTheirCanopies2017]

@@

Forest canopies moderate relative humidity in below-canopy microclimates along with temperature. In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate. With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012a]

RH, in combination with temperature, determines vapor pressure deficit, $VPD$
- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018, Tymen et al. 2017]
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
-	Higher in understory, particularly at dusk (Koike et al. 2001).
-	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
-Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

@@

**The strength of this buffering varies across forests, being influenced by both forest characteristics and the biophysical environment.**
First, buffering increases with canopy cover.
* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015]
*Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]

@@

Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b]. Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@niinemetsWorldwideAnalysisWithincanopy2015].
Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregular and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al. 2020, REFS)

@@

Photosynthesis increases with increasing temperature and saturates at quatum flux density of 400 to 700nm $m^{-2}$ $s^{-1}$, beyond this the leaf experiences heat stress; non-photochemical quenching, photoinhibition and leaf necrosis. Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves, especially in conditions of water stress or drought. [@niinemetsWorldwideAnalysisWithincanopy2015]

**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5; *revisit this!*) (Campbell and Norman 1998).
This observed by earlier studies suggested that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of Topt, Vmax anD Jmax along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photoshythesis in sun and shade leaves.
For 3 species in Panama, [@slotPhotosyntheticHeatTolerance2019] found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant. Similiar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
Mau et al. found no trend along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
